Type-Safe Client-Server Communication in Web Applications

### 2014-02-27 paper.tex
Add general description of our solution plus Hello Server example.

window.onload = function() {
  var logbox = document.getElementById('log');
  var msgbox = document.getElementById('message');
  var sock = new WebSocket('ws://example.com:24601');

@@

\section{A seamless programming model}
\section{Implementation}
\section{The Haste compiler}

@@

This is easily remedied however, by a slight compiler modification. Instead of
using \lstinline!main :: IO ()! as the program entry point, introduce a function
\lstinline!appMain :: App Done! for this purpose, and let the runtime system
start the application by invoking \lstinline!runApp config appMain!, where
\lstinline!config! contains the URL and port configuration for the server, given
by the programmer using a compiler flag. As this modification only affects the
program's entry point, which is already given ``special'' treatment in Haskell
compilers, we believe this to be a practical, non-intrusive way to remove this
limitation, and in fact made this modification to the Haste compiler for our
implementation.

>>>>

window.onload = function() {
  var logbox = document.getElementById('log');
  var msgbox = document.getElementById('message');
  var sock = new WebSocket('ws://example.com);

@@

\section{A seamless programming model}

There are many conveivable improvements to the mainstream web development model
described in the previous section. We propose an alternative programming model
based on Haskell, in which web applications are written as monolithic entities
rather than as two independent parts that just so happen to talk to each other.

Our proposed model, dubbed ``Haste.App'' in an ingenious attempt at
capitalizing on the recent popularity of the ``app'' buzzword, has the
following properties:

\begin{itemize}
  \item The programming model is synchronous, giving the programmer a simple,
        linear view of the program flow, eliminating the need to program with
        callbacks and continuations.
  \item Side-effecting code is explicitly designated to run on either the
        client or the server using the type system while pure code can be
        shared by both. Additionally, both client and server code may lift
        general IO computations, allowing for safe IO code reuse within the
        confines of the client or server designated code sections.
  \item Client-server network communication is handled through statically typed
        RPC function calls, extending the reach of Haskell's type checker over
        the network and giving the programmer advance warning when she uses
        network services incorrectly or forgets to update communication code
        as the application's internal protocol changes.
  \item The implementation is built as a library on top of the GHC and Haste
        Haskell compilers, requiring little to no compiler support.
\end{itemize}

\paragraph{A first example} While explaining of the properties of our solution
is all well and good, nothing compares to a good old Hello World example to
convey the idea. We begin by implementing a function which prints a greeting
to the server's console.

\begin{code}
import Haste.App

helloServer :: String -> Server ()
helloServer name =
  liftIO $ putStrLn (name ++ " says hello!")
\end{code}

Computations exclusive to the server side live in the \lstinline!Server! monad.
This is basically an IO monad with a few extra operations for session handling;
its main purpose is to prevent the programmer from accidentally attempting to
perform client-exclusive operations, such as popping up a browser dialog box,
on the server.

Next, we need to make the \lstinline!helloServer! function available as an RPC
function and call it from the client.

\begin{code}
appMain :: App Done
appMain = do
  greetings <- export helloServer

  runClient $ do
    name <- prompt "Hi there, what is your name?"
    onServer (greetings <.> name)
\end{code}

The \lstinline!export! function takes an arbitrary function, provided that all
its arguments as well as its return value are serializable through the aptly
named \lstinline!Serializable! type class, and returns a typed ``handle'' of
sorts. In this example, the type of \lstinline!greetings! is
\lstinline!Export (String -> Server ())!, indicating that the handle identifies
an exported function with a single \lstinline!String! argument and no return
value; exported functions all live in the \lstinline!Server! monad.

In order to make an RPC call using this handle, we must supply it with an
argument. This is done using the \lstinline!<.>! operator. Upon closer
inspection, the attentive reader may notice that its type,
\linebreak
\lstinline!Serialize a => Export (a -> b) -> a -> Export b!,
is very similar to the type of the \lstinline!<*>! operator over applicative
functors. This is not a coincidence; the \lstinline!<.>! performs the same role
for the \lstinline!Export! type as \lstinline!<*>! performs for applicative
functors. The reason for using a separate operator for this instead of making
\lstinline!Export! an instance of \lstinline!Applicative! is that since
functions embedded in the \lstinline!Export! type exist only to be called over
a network, such functions must only be applied to arguments which can be
serialized and sent over a network connection.

After applying the value obtained by \lstinline!prompt!ing the user for her
name to the exported function, we apply the \lstinline!onServer! function to
the result, which dispatches the RPC call to the server. \lstinline!onServer!
will then block until the RPC call returns.

Finally, we need to launch our application, and inform the client part of the
whereabouts of its server counterpart; in this example the server will listen
on port 1111, and since we will only run this application locally the client
should attempt to contact the server at our local machine.

\begin{code}
main :: IO ()
main =
  runApp (mkConfig "ws://localhost:1111" 1111) appMain
\end{code}

This part is entirely straightforward; the URL of the server is provided for
the benefit of the client and the port to listen on is provided for the
benefit of the server. The final argument is the main function of our
application. Figure \ref{lst:hello-server} gives the code of this example in
its entirety.

\begin{listingfloat}
\begin{code}
import Haste.App

helloServer :: String -> Server ()
helloServer name =
  liftIO $ putStrLn (name ++ " says hello!")

appMain :: App Done
appMain = do
  greetings <- export helloServer

  runClient $ do
    name <- prompt "Hi there, what is your name?"
    onServer (greetings <.> name)

main :: IO ()
main =
  runApp (mkConfig "ws://localhost:1111" 1111) appMain
\end{code}
\caption{A seamless programming model: Hello Server}
\label{lst:hello-server}
\end{listingfloat}

\section{Implementation}
\section{The Haste compiler}

@@

This is easily remedied however, by a slight compiler modification. Instead of
using \lstinline!main :: IO ()! as the program entry point, introduce a function
\lstinline!appMain :: App Done! for this purpose, and let the runtime system
start the application by invoking \lstinline!runApp config appMain!, where
\lstinline!config! contains the URL and port configuration for the server, given
by the programmer using a compiler flag. As this modification only affects the
program's entry point, which is already given ``special'' treatment in Haskell
compilers, and may even be implemented using the plugin mechanism of GHC,
requiring no modification of the compiler code at all, we believe this to be a
practical, non-intrusive way to remove this limitation, and are planning on
including this modification in our implementation.
